Documenting Social, Geographic, and Economic Inequities in COVID-19 Mortality at the County Level in the US Using Generalized Additive Models

### 2022-11-14 manuscript/manuscript.rmd
Checkin; drafting more intro

---
title: Analyzing COVID-19 Mortality at the County Level in the U.S. Using Generalized Additive
  Models
author: "Christian Testa"
date: '`r Sys.Date()`'
output: pdf_document
abstract: this is an abstract
---

```{r setup, echo = FALSE}
library(here)

@@

As we enter the third winter with novel coronavirus disease (COVID-19) in the
United States (U.S.), evidence reflecting intense disparities in COVID-19 mortality rates 
among socially advantaged and disadvantaged populations persists.  Eliminating
inequities in health outcomes is a major policy goal of the Biden
administration, representing an revitalized focus on health justice
and underlining the importance of adequate data reporting systems that measure
the prevalent health inequities.  To this end, generalized additive models
(GAMs) provide a flexible regression framework which can illustrate the evolving
roles and relationships social conditions have with trends in COVID-19
mortality.

# Background

Having passed over 1 million COVID-19 deaths in the United States earlier this
year, and facing uncertain prospects for the third COVID-19 winter looming even
as new iterations on the COVID-19 vaccines become available, it remains critical
that inequities in COVID-19 outcomes are documented and analyzed to reckon with
the unjust and unfair burden of preventable illness.  Even though the first
vaccines were granted emergency use authorization by the U.S. Food and Drug
Administration in 2020, with the first shots going in arms in December 2020,
COVID-19 is still continuing to cause hundreds of deaths a day in the U.S. in
the late summer and early fall of 2022.  The new bivalent vaccines released at
the end of August 2022 contain messenger RNA components of the original strain
as well as the BA.4 and BA.5 lineages in an effort to make the nation's immunity
more up-to-date and robust against the myriad of phylogenetic directions the
COVID-19 virus is exploring.  Despite the updated bivalent boosters representing
a significant step forward in prevention strategy, less than 4% of eligible
Americans had taken the booster in the first month after it became available. As
such, and with an enduring history of inequities in health care availability, it
is not clear that all communities will be equally able to take advantage of the
new vaccines and inequities in COVID-19 illness and mortality may persist
despite the technological innovations in vaccine therapy.

```{r figure 1, fig.cap = "Monthly estimates of COVID-19 mortality per 100,000 residents by state are shown.  States are in distinct colors to make each state's trajectory visible. Reference the online version TODO: XXX"}
knitr::include_graphics(here("analysis/01_state_trajectories/state_covid19_mortality_trajectories.png"))
```

@@

  * Counts of deaths by month with underlying cause of death as
  COVID-19 (ICD-10 code U07.1) from CDC WONDER, Provisional Multiple Cause of Death
  Data
  * Population Size Estimates for 2020 from the U.S. Census (Redistricting 
  File Public-Law 94-171 Dataset). *Note:*  While the complete data for the 2020
  U.S. Census have not yet been released, the summary tables including
  population counts have been released as mandated by Congress for the purposes
  of redistricting and updating the boundaries of the 435 voting Congressional
  Districts, which are what is made us of here as population size estimates.
  * Percent Black, percent Hispanic, percent Asian, percent American Indian or
  Alaska Native, Percent below the federal poverty line, median household
  income, median age, percent 65 years or older, and percent 75 years or older
  from the 2014-2019 5-year American Community Survey. 
  *Note:* while more recent American Community Survey 5-year estimates are
  available, survey data collection in 2020 was severely affected by the
  pandemic with the 2020 data release denoted as "experimental".

@@

@@

@@

@@

# Supplementary Materials

>>>>

---
title: Documenting Social, Geographic, and Economic Inequities in COVID-19 Mortality at the County Level in the US Using Generalized Additive Models
author: 
  - name: "Christian Testa"
    institute: "1"
    email: "ctesta@hsph.harvard.edu"
    correspondence: true
institute: 
  - "1": "Department of Social and Behavioral Sciences, Harvard T.H. Chan School of Public Health"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
abstract: |
  We present three types of applications of generalized additive models (GAMs) to
  COVID-19 mortality rates in the US for the purpose of advancing methods to
  document inequities with respect to which communities suffered disproportionate
  COVID-19 mortality rates. First, GAMs can be used to describe the changing
  relationship between COVID-19 mortality and county-level covariates
  (sociodemographic, economic, and political metrics) over time. Second, GAMs can
  be used to perform spatiotemporal smoothing that pools information over time and
  space so that county level mortality rates that tend to be noisy due to small
  population counts or stochasticity can be summarized by a smooth, dynamic latent
  surface describing the mortality risk associated with geographic locations over
  time. Third, estimation of the associations between county-level covariates and
  COVID-19 mortality after controlling for spatiotemporal effects allow for the
  distinguishing of what patterns in COVID-19 mortality were more plausibly due to
  geography than variation in county-level covariates. Each of these approaches
  provides a valuable approach and perspective to documenting inequities in
  COVID-19 mortality by addressing the question of which populations have suffered
  the worst burden of COVID-19 mortality taking into account the nonlinear
  spatial, temporal, and social patterning of disease.
  \newline

  \noindent Abbreviations used: United States (US), Coronavirus Disease 2019 (COVID-19), Generalized
  Additive Model (GAM), Centers for Disease Control and Prevention (CDC)
bibliography: bibliography.bib
csl: apa.csl
link-citations: true
urlcolor: blue
linkcolor: black
---

<!--
TODO:

Throughout/Overarching: 
  - Do I need to update all the models to separately control for age? Probably.
  - Do I want to have a supplementary materials section?
  - Talk about ecosocial theory (?) (Who benefits from harm done to whom?)

Abstract:
  - Add link to GitHub repository

Intro: 
  - add geographic background, like where the first cases were reported, 
  how COVID exploded in NYC, then the surges in the Midwest and South
  - add a primer on the relevance of spatiotemporal modeling

Methods: 
  - add intro to ICEraceinc (methods)
  - describe negative binomial models

Results:
  - add figure for temporal two-way interaction effects
  - present rate ratios (results)
  - present model comparisons (results, F-tests, AIC)

Discussion:
  - talk about how COVID has not been a win for health equity (discussion)
  - talk about limitations to being causal
  - talk about future directions

Conclusion:
  - summarize
-->

```{r setup, echo = FALSE}
suppressMessages(library(here))

@@

As we enter the third winter with the novel coronavirus disease 
COVID-19 in the United States, evidence documenting the intense disparities in
COVID-19 mortality rates comparing socially advantaged and disadvantaged
populations continues to mount. Eliminating inequities in health outcomes has
been stated as a major policy goal of the Biden administration
[@the_white_house_executive_2021; @the_white_house_advancing_2022], representing
a revitalized focus on health justice and underlining the importance of adequate
data reporting systems that measure the prevalent health inequities. To this
end, we use generalized additive models (GAMs) as a flexible regression
framework to illustrate the evolving roles and relationships sociodemographic,
geographic, and economic conditions have with respect to trends in COVID-19
mortality. The code, data, and documentation necessary to reproduce the analyses
contained in this paper are online and free to access at <https://github.com/ctesta01/covid.gradient.estimation>.

### Background

Having passed over 1 million COVID-19 deaths in the United States earlier this
year [@donovan_us_2022], and facing uncertain prospects for the third COVID-19
winter looming even as new iterations on the COVID-19 vaccines become available,
it remains critical that inequities in COVID-19 outcomes are documented and
analyzed to reckon with the unjust and unfair burden of preventable illness.
Even though the first vaccines were granted emergency use authorization by the
U.S. Food and Drug Administration in 2020 [@mayo_clinic_history_2022], with the
first shots going in arms in December 2020, COVID-19 is still continuing to
cause hundreds of deaths a day in the US in the fall of 2022
[@noauthor_united_2022].  The new bivalent vaccines released at the end of
August 2022 contain mRNA sequences from both the original strain as well as
the recently emergent BA.4 and BA.5 lineages in an effort to make the nation's immunity more
up-to-date and robust against the myriad of phylogenetic directions the COVID-19
virus is evolving to explore [@office_of_the_commissioner_coronavirus_2022].
Despite the updated bivalent boosters representing a significant step forward in
prevention strategy, less than 4% of eligible Americans had taken the booster in
the first month after it became available [@bendix_less_2022; @lambert_most_2022]. 
As such, and with an enduring history of inequities in health care access 
in the US [@ortega_ending_2021; @blendon_inequities_2002; @bailey_how_2021;
@okonkwo_covid-19_2021; @rapp_statelevel_2022;
@chrisler_ageism_2016; @feldman_health_2021; @carpenter_health_2021;
@whitehead_outness_2016], it is clear that without further intervention not all
communities will be equally able to benefit from the updated vaccines and
inequities in COVID-19 illness and mortality may persist despite the
technological innovations in vaccine technology.

### The Role of Geography in COVID-19

```{r figure-county-trends, fig.cap = "Estimates of monthly COVID-19 mortality rates per 100,000 person-years by county organized by Census Division.  For each division, the median trendline and quantile ranges are shown weighted by county population size.", out.width = "6in", fig.align = 'center', fig.env = "figure*"}
knitr::include_graphics(here("analysis/02_plot_county_trends/county_trends_by_division.png"))
```

Prior literature has demonstrated that geography has played a significant role
in the spread and impact of of COVID-19. Methods employed to highlight the role
geography plays have included quantile regression [@sigler_socio-spatial_2021],
Besag-York-Molli\'e mixed models [@whittle_ecological_2020], spatial cluster analysis
[@sugg_mapping_2021], geographically weighted regression [@mollalo_gis-based_2020; @park_covid-19_2021], and others.

In the US context, one of the key aspects to the geographic story of COVID-19's
spread and diffusion was the early surge of cases and epicenter in New York City
during March 2020 [@thompson_covid-19_2020] followed by subsequent waves of cases in the
South and Midwest [@shumaker_covid-19_2020; @scott_these_2020; @glenza_covid_2020]. As Park et al. 
stated in their article *COVID-19 Deaths in the United States: Shifts in Hot Spots over the Three Phases of the Pandemic and the Spatiotemporally Varying Impact of Pandemic Vulnerability* where they summarized
US trends from March 2020 to May 2021, "hot spots have shifted from densely populated cities and the states with a high percentage of socially vulnerable individuals to the states with relatively relaxed social distancing requirements, and then to the states with low vaccination rates" [@park_covid-19_2021].

When considering the drivers of the COVID-19 pandemic, it's necessary to note that 
geography and social conditions are inextricably linked. In July 2021, the CDC
reported that "the COVID-19 cumulative death rate in non-metropolitan areas has
exceeded that of metropolitan areas since December 2020," noting that of the
approximately 1/5th of Americans who live in rural areas, many "are considered
highly vulnerable according to CDC’s Social Vulnerability Index (SVI), which
includes factors such as housing, transportation, socioeconomic status, race,
and ethnicity" [@cdc_location_2021]. Moreover, rural communities often have
lower health insurance rates, higher disability rates, older populations, and
limited access to health care. One of CDC's Morbidity and Mortality Weekly
Reports found that vaccination against COVID-19 was lower in rural communities
than in urban communities between December 2020 and April 2021
[@murthy_disparities_2021].

Figure \@ref(fig:figure-county-trends) shows the monthly COVID-19 mortality
rates for the counties grouped within each of the nine U.S. Census Divisions
[@us_department_of_commerce_economics_and_statistics_administration_census_2000]. 
The figure summarizes each division's median mortality rates
weighted by county population size. Quantile ranges are included to illustrate
the range of mortality rates observed, again weighted by county population size.
Weighting based on county population size is used as opposed to equal weighting
for calculating the median and quantile ranges so that in divisions where low
population counties predominate, mortality rate estimates that are noisy due to
small population counts are not weighted equally to more reliable mortality rate
estimates from larger population counties. Notably, the mortality associated
with the early surge of cases starting in NYC and spreading through New York,
New Jersey, and Massachusetts is visible in the Middle Atlantic and New England
division figures. The figure also illustrates how the first peak in the
mortality time-series for states in the Midwest (West North Central, East North
Central) occurred later, in late 2020 and going into early 2021.

### The Social Determinants of COVID-19 Mortality

<!-- 
Want to cite: 
* Mude et al (racial disparities meta-analysis);
* Chen & Krieger on their article "Revealing the Unequal Burden" 
* Hanage et al COVID-19: US federal accountability for entry, spread and inequities
* Kim et al on COVID-19 magnifying the effect of health disparities (esp. re: testing (re: care providers, transportation equity))
* Romano et al (CDC MMWR on trends and disparities in hospitalizations by race/ethnicity and region)
* Okonkwo et al: accelerating health inequities
* Del Rios: COVID-19 is an inverse equity story, not a racial equity success story
* Goldstein/NYT: During the Omicron surge, Black New Yorkers were hospitalized at a rate more than twice that of white residents.
* Mueller/NYT In Rural America, Covid Hits Black and Hispanic People Hardest
* Hill et al / KFF COVID-19 Cases and Deaths by Race/Ethnicity: Current Data and Changes Over Time
* Lo et al The polarization of politics and public opinion and their effects on racial inequality in COVID mortality
* Johnson and Keating: Whites now more likely to die from covid than Blacks: Why the pandemic shifted 
* Kaashoek The Evolving Roles of US Political Partisanship and Social Vulnerability in the COVID-19 Pandemic from February 2020 - February 2021
* 
* Krieger et al on a warning against static measures

Outline:

Disparities became immediately obvious: 
  Though many have repeated the oft-quoted line "viruses do not discriminate,"
  it was immediately clear that COVID-19 would lay bare the sharp inequities
  that exist in tge US; almost immediately some groups were shown to be
  experiencing higher rates of COVID-19 infection and mortality. Those groups
  have been shown to include marginalized racial/ethnic groups, those who do not
  speak English, transportation insecurity, the elderly, those exposed to
  chronic pollution, those living in crowded housing or areas with financial
  insecurity, and those with disabilities, housing instability and/or
  homelessness, pre-existing conditions (including chronic obstructive pulmonary
  disease, diabetes, heart diseases, etc.).  Smoking.  Obesity.  Acute Kidney
  Injury.

There are some aspects of this story that we still don't know: 
  The details on COVID-19 rates are not availability stratified by both disability
  and race/ethnicity simultaneously making it difficult to understand how much of
  racial/ethnic disparities in COVID-19 are attributable to increased disability
  rates among historically (and even now) marginalized racial/ethnic groups.

  Moreover, it is known that many of the risk factors for increased COVID-19 severity 
  are prevalent at increased rates among sexual minority persons compared to heterosexual
  persons, suggesting that sexual minority persons may be at increased risk of 
  severe COVID-19.

Vaccines didn't right the boat: 
  Again, when vaccines became available, I think many thought "surely this will
  be the silver bullet that solves the problem";  inequities in who were able to
  get vaccine appointments reinforced existing patterns in internet access,
  technological literacy and availability of transportation. Even after the
  initial vaccine rollout, inequities persist in vaccination clinic locations
  with 'vaccine deserts' being more likely to be located in less affluent, less 
  white communities.

Throughout this timeline, it has become increasingly apparent that local-level
politics represents a critical factor influencing the trajectory of COVID-19.

COVID is not a racial equity success story: 
  Finally, racial inequities in COVID-19 in the US have been widening and
  shrinking during the course of the pandemic, largely widening during surges
  and shrinking when case rates and mortality have been lower.  During Fall 
  2022, the age-adjusted COVID-19 mortality rate for white non-Hispanic 
  people surpassed the mortality rate for Black or African/American people. 
  This has been explained as being due to those who still remain unvaccinated
  being more likely to be Republican and white. While reducing racial inequities
  in COVID-19 mortality rates during Fall 2022, may sound like a good thing, the
  cumulative burden of COVID-19 leaving in its wake long-COVID, lost loved ones,
  jobs lost, and damage to individuals' health that may never be reversed that
  disproportionately falls on marginalized racial/ethnic groups does not
  represent a health equity success story.
-->

  - Insert section on prior modeling results
  - Insert discussion motivating the need for spatiotemporal modeling

@@

  * Counts of COVID-19 deaths [@the_new_york_times_coronavirus_2021].
  * Population size estimates for 2020 from the U.S. Census (Redistricting 
  File Public-Law 94-171 Dataset) [@us_census_bureau_2020_2021].
  * Median age, median household income, racial/ethnic composition, population
  density, percent below the federal poverty line, and number of households with
  high (\$100k+)/low (<\$25k) household income by racial/ethnic group from the
  2014-2019 5-year American Community Survey [@us_census_bureau_american_2020]
  through the `tidyverse` R package [@walker_tidycensus_2022]. 
  * Votes cast in the 2020 presidential election [@mit_county_2022]

@@

Traditional generalized linear models fit a regression model 
using the functional relation $g(\mu_i) = \mathbf X_i \beta$ 
where 
Examples of common link functions include the identity function for linear regression,
$\log$ for Poisson regression, and the $\text{logit}$ function for logistic regression.

Generalized additive models (GAMs) improve upon generalized linear models by
allowing for the fitting of smooth functions that transform the $x$ variables.
This is a convenient means to include nonlinear relationships between the
outcome and predictor variables. Whereas a generalized linear model may have looked like 
$$ g(\mu_i) = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i3} ...$$ 
a generalized additive model could look like 
$$ g(\mu_i) = \mathbf A_i \mathbf \theta + f_1(x_{i1}) + f_2(x_{i2}) + f_3(x_{i3}, x_{i4}) + \dots$$
where $\mu_i \equiv \mathbb E(Y_i)$, $\mathbf A_i$ is a row of the model matrix 
for any strictly non-parametric model components, $\mathbf \theta$ is the corresponding
parameter vector, and the $f_j$ are smooth functions of the covariates $x_k$ [@wood_generalized_2017].

@@

## Evaluating Spatial and Temporal Autocorrelation

```{r figure county neighboring graph, fig.cap = "A) The spatial neighbor graph for counties in the US. In this graph, county centroids are connected if the two counties are adjacent to one another (i.e., share a boundary). This and similar graphs that represent neighbor relationships or geographic proximity are often used in modeling contexts to create spatial weights that account for which regions help to explain a given region's observed data.", fig.show='hold', out.width = '50%', fig.env = "figure*"} 
desired_caption <- 
  " B) Moran's scatter plot \\citep{anselin_simple_1996} depicts for each region (in our case, counties) on the x-axis the region rate and on the y-axis the average of the region's neighbors' rates. This diagnostic plot provides a simple means to assess to what degree neighboring regions share similar rates. In a setting where regions have high correlation, Moran's scatter plot will exhibit a high degree of positive correlation."
knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/county_neighboring_relations.png")) 
knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/morans_plot.png"))
```

```{r figure 4, fig.cap = ""}
knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/temporal_acf.png"))
```

## Regression Modeling Results

@@

```{r figure 5, fig.cap = ""}
knitr::include_graphics(here("analysis/09_spatiotemporal_models/temporal_panel_figure.png"))
```

```{r figure 6, fig.cap = ""}
knitr::include_graphics(here("analysis/09_spatiotemporal_models/panel_figure.png"))
```

@@
